---
title: "main"
format: html
editor: visual
---

## Packages

```{r}
library(dplyr)
library(readr)
library(tidyverse)
library(monomvn)
library(BoomSpikeSlab)
library(car)
library(MASS)
library(ggplot2)
library(MCMCpack)
```

## Data

```{r}
FOOD_DATA_GROUP1 <- read_csv("FINAL FOOD DATASET/FOOD-DATA-GROUP1.csv")
FOOD_DATA_GROUP2 <- read_csv("FINAL FOOD DATASET/FOOD-DATA-GROUP2.csv")
FOOD_DATA_GROUP3 <- read_csv("FINAL FOOD DATASET/FOOD-DATA-GROUP3.csv")
FOOD_DATA_GROUP4 <- read_csv("FINAL FOOD DATASET/FOOD-DATA-GROUP4.csv")
FOOD_DATA_GROUP5 <- read_csv("FINAL FOOD DATASET/FOOD-DATA-GROUP5.csv")

FOOD_DATA <- bind_rows(
  dplyr::mutate(FOOD_DATA_GROUP1, group = "group1"),
  dplyr::mutate(FOOD_DATA_GROUP2, group = "group2"),
  dplyr::mutate(FOOD_DATA_GROUP3, group = "group3"),
  dplyr::mutate(FOOD_DATA_GROUP4, group = "group4"),
  dplyr::mutate(FOOD_DATA_GROUP5, group = "group5")
) %>%
  # remove first two columns
  dplyr::select(-1, -2) %>%
  # replace spaces with underscores in colnames
  dplyr::rename_with(~ gsub(" ", "_", .x))

df <- FOOD_DATA %>%
  dplyr::select(-food, -group, -Nutrition_Density) %>%
  drop_na()

y <- df$Caloric_Value
X <- df %>% dplyr::select(-Caloric_Value)

X_scaled <- scale(X)

# Fat multicollinearity
plot(unlist(FOOD_DATA[, "Saturated_Fats"] + FOOD_DATA[, "Monounsaturated_Fats"] + FOOD_DATA[, "Polyunsaturated_Fats"]),
     unlist(FOOD_DATA[, "Fat"]))
abline(0, 1)


# ggplot version
FOOD_DATA %>%
  mutate(
    Sum_Fats = Saturated_Fats + Monounsaturated_Fats + Polyunsaturated_Fats
  ) %>%
  ggplot(aes(x = Sum_Fats, y = Fat)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = "dashed") +
  labs(
    x = "Saturated + Mono + Poly Fats",
    y = "Total Fat",
    title = "Fat Multicollinearity Check"
  ) +
  theme_minimal()
```

## EDA

```{r}
# Density Plot for the Response
df %>% 
  filter(Caloric_Value <= quantile(Caloric_Value, 0.975)) %>% 
  ggplot(aes(x = Caloric_Value)) +
  geom_density(alpha = 0.5, fill = "skyblue") +
  labs(
    x = "Caloric Value",
    y = "Density"
  )

# Density Plots for some regressors
some_regressors <- c(
  "Fat", "Carbohydrates", "Protein",
  "Saturated_Fats", "Monounsaturated_Fats", "Polyunsaturated_Fats",
  "Sugars", "Vitamin_A", "Vitamin_E",
  "Water", "Cholesterol", "Magnesium")

df_long <- df[c(some_regressors, "Caloric_Value")] %>%
  filter(if_all(all_of(some_regressors), ~ . <= quantile(., 0.975))) %>% # . >= quantile(., 0.025) &
  pivot_longer(
    cols = all_of(some_regressors), # or cols = c(A, B, C)
    names_to = "variable",
    values_to = "value"
  )

ggplot(df_long, aes(x = value)) +
  geom_density(alpha = 0.5, fill = "skyblue") +
  facet_wrap(~ variable, scales = "free") + # Creates separate plots for each variable
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Regressor Value",
    y = "Density"
  )

# Correlation Plot for some regressors
library(corrplot)

M = cor(df[c("Caloric_Value", "Fat", "Saturated_Fats", "Monounsaturated_Fats",
             "Polyunsaturated_Fats", "Carbohydrates", "Sugars", "Protein")])
colnames(M) <- c("Caloric Value", "Fats", "Saturated Fats", "Monounsaturated Fats",
                 "Polyunsaturated Fats", "Carbohydrates", "Sugars", "Protein")
rownames(M) <- colnames(M)
corrplot.mixed(M, diag="l", tl.pos = "lt", tl.col = "#000011", tl.srt = 45)

# Regressors Vs Response
num_labels <- c(
  "Caloric_Value" = "Caloric Value",
  "Saturated_Fats" = "Saturated Fats",
  "Monounsaturated_Fats" = "Monounsaturated Fats",
  "Polyunsaturated_Fats" = "Polyunsaturated Fats",
  "Fat" = "Fat",
  "Carbohydrates" = "Carbohydrates",
  "Sugars" = "Sugars",
  "Protein" = "Protein"
)

main_regressors <- c(
  "Fat", "Carbohydrates", "Protein",
  "Saturated_Fats", "Monounsaturated_Fats", "Polyunsaturated_Fats")

df_long <- df[c(main_regressors, "Caloric_Value")] %>%
  pivot_longer(
    cols = all_of(main_regressors),
    names_to = "regressor_variable",
    values_to = "regressor_value"
  )


ggplot(df_long, aes(x = regressor_value, y = Caloric_Value)) +
  geom_point(alpha = 0.5, color = "skyblue") +
  facet_wrap(~ factor(regressor_variable, 
                      levels = main_regressors), 
             scales = "free",
             labeller = as_labeller(num_labels)) +
  labs(
    x = "Regressor Value",
    y = "Caloric Value (Response)") +
  theme_minimal()
```

## Variable Selection

### 1. Bayesian Lasso

```{r}
set.seed(123)
bl_fit <- blasso(
  X = X_scaled,
  y = y,
  T = 10000,
  thin = 5,
  verb = 0
)

# Posterior mean of each coefficient
beta_mean <- apply(bl_fit$beta, 2, mean)
names(beta_mean) <- colnames(X_scaled)

# Rank nutrients by absolute effect size
ranked_lasso <- sort(abs(beta_mean), decreasing = TRUE)
head(ranked_lasso, 20)

beta_ci <- t(apply(bl_fit$beta, 2, quantile, probs = c(0.025, 0.975)))
colnames(beta_ci) <- c("low", "high")

# Selected nutrients from Bayesian Lasso
selected_lasso_ci <- rownames(beta_ci)[beta_ci[, "low"] > 0 | beta_ci[, "high"] < 0]
selected_lasso_ci # The Bayesian lasso thinks those predictors have credible, non-zero effects on "Caloric Value".

idx_sel <- as.integer(sub("b\\.", "", selected_lasso_ci))
idx_sel 

selected_nutrients <- colnames(X_scaled)[idx_sel]
selected_nutrients

```

### 2. Bayesian Lasso Without Fat

```{r}
X_without_fat <- X %>% dplyr::select(-Fat)

X_scaled_without_fat <- scale(X_without_fat)

# Bayesian Lasso without fat
set.seed(123)
bl_fit_without_fat <- blasso(
  X = X_scaled_without_fat,
  y = y,
  T = 10000,
  thin = 5,
  verb = 0
)

# Posterior mean of each coefficient
beta_mean_without_fat <- apply(bl_fit_without_fat$beta, 2, mean)
names(beta_mean_without_fat) <- colnames(X_scaled_without_fat)

# Rank nutrients by absolute effect size
ranked_lasso_without_fat <- sort(abs(beta_mean_without_fat), decreasing = TRUE)
head(ranked_lasso_without_fat, 20)

beta_ci_without_fat <- t(apply(bl_fit_without_fat$beta, 2, quantile, probs = c(0.025, 0.975)))
colnames(beta_ci_without_fat) <- c("low", "high")

# Selected nutrients from Bayesian Lasso
selected_lasso_ci_without_fat <- rownames(beta_ci_without_fat)[beta_ci_without_fat[, "low"] > 0 | beta_ci_without_fat[, "high"] < 0]
selected_lasso_ci_without_fat # The Bayesian lasso thinks those predictors have credible, non-zero effects on "Caloric Value".

idx_sel_without_fat <- as.integer(sub("b\\.", "", selected_lasso_ci_without_fat))
idx_sel_without_fat 

selected_nutrients_without_fat <- colnames(X_scaled_without_fat)[idx_sel_without_fat]
selected_nutrients_without_fat
```

### 3. Bayesian Lasso Without Fat Types

```{r}
X_without_fat_types <- X %>% dplyr::select(-c(Saturated_Fats, Monounsaturated_Fats, Polyunsaturated_Fats))
X_scaled_without_fat_types <- scale(X_without_fat_types)

set.seed(123)
bl_fit_without_fat_types <- blasso(
  X = X_scaled_without_fat_types,
  y = y,
  T = 10000,
  thin = 5,
  verb = 0
)

# Posterior mean of each coefficient
beta_mean_without_fat_types <- apply(bl_fit_without_fat_types$beta, 2, mean)
names(beta_mean_without_fat_types) <- colnames(X_scaled_without_fat_types)

# Rank nutrients by absolute effect size
ranked_lasso_without_fat_types <- sort(abs(beta_mean_without_fat_types), decreasing = TRUE)
head(ranked_lasso_without_fat_types, 20)

# 95% credible intervals
beta_ci_without_fat_types <- t(apply(
  bl_fit_without_fat_types$beta,
  2,
  quantile,
  probs = c(0.025, 0.975)
))
colnames(beta_ci_without_fat_types) <- c("low", "high")

# Selected nutrients (credible interval excludes zero)
selected_lasso_ci_without_fat_types <- rownames(beta_ci_without_fat_types)[
  beta_ci_without_fat_types[, "low"] > 0 |
    beta_ci_without_fat_types[, "high"] < 0
]

selected_lasso_ci_without_fat_types   # Variables with credible non-zero effects

# Extract original nutrient names
idx_sel_without_fat_types <- as.integer(sub("b\\.", "", selected_lasso_ci_without_fat_types))
idx_sel_without_fat_types

selected_nutrients_without_fat_types <- colnames(X_scaled_without_fat_types)[idx_sel_without_fat_types]
selected_nutrients_without_fat_types
```

### 4. Spike and Slab Prior

```{r}
set.seed(123)
df_model <- data.frame(
  y = as.numeric(y),
  as.data.frame(X_scaled)
)

ss_fit <- lm.spike(
  y ~ .,
  data  = df_model,
  niter = 6000,                 
  expected.model.size = 10    # prior belief of how many predictors matter
)

burn_in <- 1000

ss_sum <- summary(ss_fit, burn = burn_in)

# matrix with columns:
# 1 post mean, 2 post sd,
# 3 mean|nonzero, 4 sd|nonzero,
# 5 Pr(nonzero)
coef_table <- ss_sum$coefficients

inc_prob <- coef_table[, 5]
inc_prob <- inc_prob[names(inc_prob) != "(Intercept)"]

inc_prob_sorted <- sort(inc_prob, decreasing = TRUE)
selected_vars <- names(inc_prob)[inc_prob > 0.5] # keep any nutrient whose coefficient is nonzero in more than half the posterior samples.
selected_vars
```

### 5. Spike and Slab Prior Without Fat

```{r}
X_without_fat <- X %>% dplyr::select(-Fat)
X_scaled_without_fat <- scale(X_without_fat)
set.seed(123)
df_model_nofat <- data.frame(
  y = as.numeric(y),
  as.data.frame(X_scaled_without_fat)
)

# Fit spike-and-slab regression
ss_fit_nofat <- lm.spike(
  y ~ .,
  data  = df_model_nofat,
  niter = 6000,
  expected.model.size = 10
)

burn_in <- 1000

ss_sum_nofat <- summary(ss_fit_nofat, burn = burn_in)

coef_table_nofat <- ss_sum_nofat$coefficients

inc_prob_nofat <- coef_table_nofat[, 5]
inc_prob_nofat <- inc_prob_nofat[names(inc_prob_nofat) != "(Intercept)"]

inc_prob_nofat_sorted <- sort(inc_prob_nofat, decreasing = TRUE)

selected_vars_nofat <- names(inc_prob_nofat)[inc_prob_nofat > 0.5]

inc_prob_nofat_sorted
selected_vars_nofat
```

### 6. Spike and Slab Prior Without Fat Types

```{r}
X_without_fat_types <- X %>% dplyr::select(-c(Saturated_Fats, Monounsaturated_Fats, Polyunsaturated_Fats))
X_scaled_without_fat_types <- scale(X_without_fat_types)
set.seed(123)
df_model_nofat_types <- data.frame(
  y = as.numeric(y),
  as.data.frame(X_scaled_without_fat_types)
)

# Fit spike-and-slab regression
ss_fit_nofat_types <- lm.spike(
  y ~ .,
  data  = df_model_nofat_types,
  niter = 6000,
  expected.model.size = 10
)

burn_in <- 1000

ss_sum_nofat_types <- summary(ss_fit_nofat_types, burn = burn_in)

coef_table_nofat_types <- ss_sum_nofat_types$coefficients

inc_prob_nofat_types <- coef_table_nofat_types[, 5]
inc_prob_nofat_types <- inc_prob_nofat_types[names(inc_prob_nofat_types) != "(Intercept)"]

inc_prob_nofat_types_sorted <- sort(inc_prob_nofat_types, decreasing = TRUE)

selected_vars_nofat_types <- names(inc_prob_nofat_types)[inc_prob_nofat_types > 0.5]

inc_prob_nofat_types_sorted
selected_vars_nofat_types
```

## Bayesian Linear Regression

### 1. Model 1: Full model with all variables as regressors

```{r}
n_full <- nrow(X_scaled)
p_full <- ncol(X_scaled)

xi_full <- rep(0, p_full)
Omega_full <- 10^3 * diag(p_full)
alpha_full <- 0.01
b_full <- 0.01  # using b instead of beta for rate parameter

Omega_inverse_full <- solve(Omega_full)
Q_beta_full <- t(as.matrix(X_scaled)) %*% X_scaled + Omega_inverse_full
Q_beta_inverse_full <- solve(Q_beta_full)
l_beta_full <- t(as.matrix(X_scaled)) %*% y + Omega_inverse_full %*% xi_full

# Sampling posterior
samples_full <- 10^4
sigma_2_samples_full <- rinvgamma(
  samples_full,
  alpha_full + n_full / 2,
  b_full + t(y) %*% y / 2 +
    t(xi_full) %*% Omega_inverse_full %*% xi_full / 2 -
    t(l_beta_full) %*% Q_beta_inverse_full %*% l_beta_full / 2
)

beta_samples_full <- matrix(NA, nrow = samples_full, ncol = p_full)
for (s in 1:samples_full) {
  beta_samples_full[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_full %*% l_beta_full,
    Sigma = sigma_2_samples_full[s] * Q_beta_inverse_full
  )
}

hist(sigma_2_samples_full, xlab = expression(sigma^2))

# Credible intervals of betas
beta_summary_full <- data.frame(
  term = colnames(X_scaled),
  mean = colMeans(beta_samples_full),
  low = apply(beta_samples_full, 2, quantile, 0.025),
  high = apply(beta_samples_full, 2, quantile, 0.975)
)
beta_summary_full

selected_betas_full <- beta_summary_full[
  beta_summary_full[, "low"] > 0 | beta_summary_full[, "high"] < 0,
]
selected_betas_full
```

### 2. Model 2: Bayesian Lasso Model

```{r}
X_scaled_BL_F <- X_scaled[, selected_nutrients, drop = FALSE]

n_BL_F <- nrow(X_scaled_BL_F)
p_BL_F <- ncol(X_scaled_BL_F)

# Prior settings
xi_BL_F <- rep(0, p_BL_F)
Omega_BL_F <- 10^3 * diag(p_BL_F)
alpha_BL_F <- 0.01
b_BL_F <- 0.01  # rate parameter

Omega_inverse_BL_F <- solve(Omega_BL_F)

Q_beta_BL_F <- t(as.matrix(X_scaled_BL_F)) %*% X_scaled_BL_F + Omega_inverse_BL_F
Q_beta_inverse_BL_F <- solve(Q_beta_BL_F)

l_beta_BL_F <- t(as.matrix(X_scaled_BL_F)) %*% y + Omega_inverse_BL_F %*% xi_BL_F

# Posterior sampling
samples_BL_F <- 10^4

sigma_2_samples_BL_F <- rinvgamma(
  samples_BL_F,
  alpha_BL_F + n_BL_F / 2,
  b_BL_F + t(y) %*% y / 2 +
    t(xi_BL_F) %*% Omega_inverse_BL_F %*% xi_BL_F / 2 -
    t(l_beta_BL_F) %*% Q_beta_inverse_BL_F %*% l_beta_BL_F / 2
)

beta_samples_BL_F <- matrix(NA, nrow = samples_BL_F, ncol = p_BL_F)

for (s in 1:samples_BL_F) {
  beta_samples_BL_F[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_BL_F %*% l_beta_BL_F,
    Sigma = sigma_2_samples_BL_F[s] * Q_beta_inverse_BL_F
  )
}

hist(sigma_2_samples_BL_F, xlab = expression(sigma^2))

# Credible intervals
beta_summary_BL_F <- data.frame(
  term = colnames(X_scaled_BL_F),
  mean = colMeans(beta_samples_BL_F),
  low = apply(beta_samples_BL_F, 2, quantile, 0.025),
  high = apply(beta_samples_BL_F, 2, quantile, 0.975)
)

beta_summary_BL_F

selected_betas_BL_F <- beta_summary_BL_F[
  beta_summary_BL_F$low > 0 | beta_summary_BL_F$high < 0,
]

selected_betas_BL_F
```

### 3. Model 3: Bayesian Lasso Model without Fat

```{r}
X_scaled_BL <- X_scaled[, selected_nutrients_without_fat, drop = FALSE]

n_BL <- nrow(X_scaled_BL)
p_BL <- ncol(X_scaled_BL)

xi_BL <- rep(0, p_BL)
Omega_BL <- 10^3 * diag(p_BL)
alpha_BL <- 0.01
b_BL <- 0.01  # rate parameter

Omega_inverse_BL <- solve(Omega_BL)

Q_beta_BL <- t(as.matrix(X_scaled_BL)) %*% X_scaled_BL + Omega_inverse_BL
Q_beta_inverse_BL <- solve(Q_beta_BL)

l_beta_BL <- t(as.matrix(X_scaled_BL)) %*% y + Omega_inverse_BL %*% xi_BL

# Posterior sampling
samples_BL <- 10^4

sigma_2_samples_BL <- rinvgamma(
  samples_BL,
  alpha_BL + n_BL / 2,
  b_BL + t(y) %*% y / 2 +
    t(xi_BL) %*% Omega_inverse_BL %*% xi_BL / 2 -
    t(l_beta_BL) %*% Q_beta_inverse_BL %*% l_beta_BL / 2
)

beta_samples_BL <- matrix(NA, nrow = samples_BL, ncol = p_BL)

for (s in 1:samples_BL) {
  beta_samples_BL[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_BL %*% l_beta_BL,
    Sigma = sigma_2_samples_BL[s] * Q_beta_inverse_BL
  )
}

# Plot sigma^2 samples
hist(sigma_2_samples_BL, xlab = expression(sigma^2))

# Credible intervals
beta_summary_BL <- data.frame(
  term = colnames(X_scaled_BL),
  mean = colMeans(beta_samples_BL),
  low = apply(beta_samples_BL, 2, quantile, 0.025),
  high = apply(beta_samples_BL, 2, quantile, 0.975)
)

beta_summary_BL

# Select coefficients with credible intervals not covering 0
selected_betas_BL <- beta_summary_BL[
  beta_summary_BL$low > 0 | beta_summary_BL$high < 0,
]

selected_betas_BL
```

### 4. Model 4: Bayesian Lasso Model without Fat Types

```{r}
X_scaled_BL_nofat_types <- X_scaled[, selected_nutrients_without_fat_types, drop = FALSE]

n_BL_nofat_types <- nrow(X_scaled_BL_nofat_types)
p_BL_nofat_types <- ncol(X_scaled_BL_nofat_types)

xi_BL_nofat_types <- rep(0, p_BL_nofat_types)
Omega_BL_nofat_types <- 10^3 * diag(p_BL_nofat_types)
alpha_BL_nofat_types <- 0.01
b_BL_nofat_types <- 0.01  # rate parameter

Omega_inverse_BL_nofat_types <- solve(Omega_BL_nofat_types)

Q_beta_BL_nofat_types <- t(as.matrix(X_scaled_BL_nofat_types)) %*% X_scaled_BL_nofat_types +
  Omega_inverse_BL_nofat_types
Q_beta_inverse_BL_nofat_types <- solve(Q_beta_BL_nofat_types)

l_beta_BL_nofat_types <- t(as.matrix(X_scaled_BL_nofat_types)) %*% y +
  Omega_inverse_BL_nofat_types %*% xi_BL_nofat_types

# Posterior sampling
samples_BL_nofat_types <- 10^4

sigma_2_samples_BL_nofat_types <- rinvgamma(
  samples_BL_nofat_types,
  alpha_BL_nofat_types + n_BL_nofat_types / 2,
  b_BL_nofat_types + t(y) %*% y / 2 +
    t(xi_BL_nofat_types) %*% Omega_inverse_BL_nofat_types %*% xi_BL_nofat_types / 2 -
    t(l_beta_BL_nofat_types) %*% Q_beta_inverse_BL_nofat_types %*% l_beta_BL_nofat_types / 2
)

beta_samples_BL_nofat_types <- matrix(NA,
                                      nrow = samples_BL_nofat_types,
                                      ncol = p_BL_nofat_types
)

for (s in 1:samples_BL_nofat_types) {
  beta_samples_BL_nofat_types[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_BL_nofat_types %*% l_beta_BL_nofat_types,
    Sigma = sigma_2_samples_BL_nofat_types[s] * Q_beta_inverse_BL_nofat_types
  )
}

# Plot sigma^2 samples
hist(sigma_2_samples_BL_nofat_types, xlab = expression(sigma^2))

# Credible intervals
beta_summary_BL_nofat_types <- data.frame(
  term = colnames(X_scaled_BL_nofat_types),
  mean = colMeans(beta_samples_BL_nofat_types),
  low = apply(beta_samples_BL_nofat_types, 2, quantile, 0.025),
  high = apply(beta_samples_BL_nofat_types, 2, quantile, 0.975)
)

beta_summary_BL_nofat_types

# Select coefficients with credible intervals not covering 0
selected_betas_BL_nofat_types <- beta_summary_BL_nofat_types[
  beta_summary_BL_nofat_types$low > 0 |
    beta_summary_BL_nofat_types$high < 0,
]

selected_betas_BL_nofat_types
```

### 5. Model 5: Spike and Slab

```{r}
X_scaled_SS <- X_scaled[, selected_vars, drop = FALSE]
# If spike & slab chose nothing, stop early
if (length(selected_vars) == 0) {
  stop("Spike&Slab selected no variables (inc_prob > 0.5). Try lowering threshold.")
}

n_SS <- nrow(X_scaled_SS)
p_SS <- ncol(X_scaled_SS)

# Prior hyperparameters (same weakly-informative N-Inv-Gamma setup)
xi_SS <- rep(0, p_SS)
Omega_SS <- 10^3 * diag(p_SS)
alpha_SS <- 0.01
b_SS <- 0.01

Omega_inverse_SS <- solve(Omega_SS)

# Posterior precision and mean terms
Q_beta_SS <- t(as.matrix(X_scaled_SS)) %*% X_scaled_SS + Omega_inverse_SS
Q_beta_inverse_SS <- solve(Q_beta_SS)

l_beta_SS <- t(as.matrix(X_scaled_SS)) %*% y + Omega_inverse_SS %*% xi_SS

# Posterior sampling
samples_SS <- 10^4

sigma_2_samples_SS <- rinvgamma(
  samples_SS,
  alpha_SS + n_SS / 2,
  b_SS + t(y) %*% y / 2 +
    t(xi_SS) %*% Omega_inverse_SS %*% xi_SS / 2 -
    t(l_beta_SS) %*% Q_beta_inverse_SS %*% l_beta_SS / 2
)

beta_samples_SS <- matrix(NA, nrow = samples_SS, ncol = p_SS)

for (s in 1:samples_SS) {
  beta_samples_SS[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_SS %*% l_beta_SS,
    Sigma = sigma_2_samples_SS[s] * Q_beta_inverse_SS
  )
}

# Plot sigma^2 posterior samples
hist(sigma_2_samples_SS, xlab = expression(sigma^2),
     main = "Posterior draws of sigma^2 (Spike&Slab Model)")

# Credible intervals for betas
beta_summary_SS <- data.frame(
  term = colnames(X_scaled_SS),
  mean = colMeans(beta_samples_SS),
  low = apply(beta_samples_SS, 2, quantile, 0.025),
  high = apply(beta_samples_SS, 2, quantile, 0.975)
)

beta_summary_SS

# Select coefficients whose 95% CI excludes 0
selected_betas_SS <- beta_summary_SS[
  beta_summary_SS$low > 0 | beta_summary_SS$high < 0,
]

selected_betas_SS
```

### 6. Model 6: Spike and Slab Prior Without Fat

```{r}
X_scaled_SS_nofat <- X_scaled[, selected_vars_nofat, drop = FALSE]

# If spike & slab chose nothing, stop early
if (length(selected_vars_nofat) == 0) {
  stop("Spike&Slab (no fat) selected no variables (inc_prob > 0.5). Try lowering threshold.")
}

n_SS_nofat <- nrow(X_scaled_SS_nofat)
p_SS_nofat <- ncol(X_scaled_SS_nofat)

# Prior hyperparameters
xi_SS_nofat <- rep(0, p_SS_nofat)
Omega_SS_nofat <- 10^3 * diag(p_SS_nofat)
alpha_SS_nofat <- 0.01
b_SS_nofat <- 0.01

Omega_inverse_SS_nofat <- solve(Omega_SS_nofat)

# Posterior precision and mean
Q_beta_SS_nofat <- t(as.matrix(X_scaled_SS_nofat)) %*% X_scaled_SS_nofat +
  Omega_inverse_SS_nofat
Q_beta_inverse_SS_nofat <- solve(Q_beta_SS_nofat)

l_beta_SS_nofat <- t(as.matrix(X_scaled_SS_nofat)) %*% y +
  Omega_inverse_SS_nofat %*% xi_SS_nofat

# Posterior sampling
samples_SS_nofat <- 10^4

sigma_2_samples_SS_nofat <- rinvgamma(
  samples_SS_nofat,
  alpha_SS_nofat + n_SS_nofat / 2,
  b_SS_nofat + t(y) %*% y / 2 +
    t(xi_SS_nofat) %*% Omega_inverse_SS_nofat %*% xi_SS_nofat / 2 -
    t(l_beta_SS_nofat) %*% Q_beta_inverse_SS_nofat %*% l_beta_SS_nofat / 2
)

beta_samples_SS_nofat <- matrix(NA,
                                nrow = samples_SS_nofat,
                                ncol = p_SS_nofat
)

for (s in 1:samples_SS_nofat) {
  beta_samples_SS_nofat[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_SS_nofat %*% l_beta_SS_nofat,
    Sigma = sigma_2_samples_SS_nofat[s] * Q_beta_inverse_SS_nofat
  )
}

# Plot sigma^2 posterior samples
hist(
  sigma_2_samples_SS_nofat,
  xlab = expression(sigma^2),
  main = "Posterior draws of sigma^2 (Spike&Slab Model, No Fat)"
)

# Credible intervals
beta_summary_SS_nofat <- data.frame(
  term = colnames(X_scaled_SS_nofat),
  mean = colMeans(beta_samples_SS_nofat),
  low = apply(beta_samples_SS_nofat, 2, quantile, 0.025),
  high = apply(beta_samples_SS_nofat, 2, quantile, 0.975)
)

beta_summary_SS_nofat

# Select coefficients with 95% CI excluding 0
selected_betas_SS_nofat <- beta_summary_SS_nofat[
  beta_summary_SS_nofat$low > 0 |
    beta_summary_SS_nofat$high < 0,
]

selected_betas_SS_nofat
```

### 7. Model 7: Spike and Slab Prior Without Fat Types

```{r}
X_scaled_SS_nofat_types <- X_scaled[, selected_vars_nofat_types, drop = FALSE]

# If spike & slab chose nothing, stop early
if (length(selected_vars_nofat_types) == 0) {
  stop("Spike&Slab (no fat types) selected no variables (inc_prob > 0.5). Try lowering threshold.")
}

n_SS_nofat_types <- nrow(X_scaled_SS_nofat_types)
p_SS_nofat_types <- ncol(X_scaled_SS_nofat_types)

# Prior hyperparameters
xi_SS_nofat_types <- rep(0, p_SS_nofat_types)
Omega_SS_nofat_types <- 10^3 * diag(p_SS_nofat_types)
alpha_SS_nofat_types <- 0.01
b_SS_nofat_types <- 0.01

Omega_inverse_SS_nofat_types <- solve(Omega_SS_nofat_types)

# Posterior precision and mean
Q_beta_SS_nofat_types <- t(as.matrix(X_scaled_SS_nofat_types)) %*% X_scaled_SS_nofat_types +
  Omega_inverse_SS_nofat_types
Q_beta_inverse_SS_nofat_types <- solve(Q_beta_SS_nofat_types)

l_beta_SS_nofat_types <- t(as.matrix(X_scaled_SS_nofat_types)) %*% y +
  Omega_inverse_SS_nofat_types %*% xi_SS_nofat_types

# Posterior sampling
samples_SS_nofat_types <- 10^4

sigma_2_samples_SS_nofat_types <- rinvgamma(
  samples_SS_nofat_types,
  alpha_SS_nofat_types + n_SS_nofat_types / 2,
  b_SS_nofat_types + t(y) %*% y / 2 +
    t(xi_SS_nofat_types) %*% Omega_inverse_SS_nofat_types %*% xi_SS_nofat_types / 2 -
    t(l_beta_SS_nofat_types) %*% Q_beta_inverse_SS_nofat_types %*% l_beta_SS_nofat_types / 2
)

beta_samples_SS_nofat_types <- matrix(
  NA,
  nrow = samples_SS_nofat_types,
  ncol = p_SS_nofat_types
)

for (s in 1:samples_SS_nofat_types) {
  beta_samples_SS_nofat_types[s, ] <- mvrnorm(
    1,
    Q_beta_inverse_SS_nofat_types %*% l_beta_SS_nofat_types,
    Sigma = sigma_2_samples_SS_nofat_types[s] * Q_beta_inverse_SS_nofat_types
  )
}

# Plot sigma^2 posterior samples
hist(
  sigma_2_samples_SS_nofat_types,
  xlab = expression(sigma^2),
  main = "Posterior draws of sigma^2 (Spike&Slab Model, No Fat Types)"
)

# Credible intervals
beta_summary_SS_nofat_types <- data.frame(
  term = colnames(X_scaled_SS_nofat_types),
  mean = colMeans(beta_samples_SS_nofat_types),
  low = apply(beta_samples_SS_nofat_types, 2, quantile, 0.025),
  high = apply(beta_samples_SS_nofat_types, 2, quantile, 0.975)
)

beta_summary_SS_nofat_types

# Select coefficients with 95% CI excluding 0
selected_betas_SS_nofat_types <- beta_summary_SS_nofat_types[
  beta_summary_SS_nofat_types$low > 0 |
    beta_summary_SS_nofat_types$high < 0,
]

selected_betas_SS_nofat_types
```

## Test and Train

### Train Test Split

```{r}
set.seed(123)

n <- nrow(df)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
test_idx  <- setdiff(seq_len(n), train_idx)

X_train <- X[train_idx, , drop = FALSE]
X_test  <- X[test_idx,  , drop = FALSE]
y_train <- y[train_idx]
y_test  <- y[test_idx]

# ---- scale using TRAIN stats only
train_means <- apply(X_train, 2, mean)
train_sds   <- apply(X_train, 2, sd)

scale_with_train <- function(Xmat, mu, sdv) {
  sweep(sweep(as.matrix(Xmat), 2, mu, "-"), 2, sdv, "/")
}

X_train_scaled <- scale_with_train(X_train, train_means, train_sds)
X_test_scaled  <- scale_with_train(X_test,  train_means, train_sds)
```

### Help Function: Conjugate Bayesian LR + Test R\^2 + Test MSE

```{r}
fit_conjugate_blr <- function(Xtr_sc, ytr, Xte_sc, yte,
                              samples = 1e4,
                              alpha0 = 0.01, b0 = 0.01,
                              Omega_scale = 1e3) {
  
  n_tr <- nrow(Xtr_sc)
  p_tr <- ncol(Xtr_sc)
  
  xi <- rep(0, p_tr)
  Omega <- Omega_scale * diag(p_tr)
  Omega_inv <- solve(Omega)
  
  Q_beta <- t(Xtr_sc) %*% Xtr_sc + Omega_inv
  Q_beta_inv <- solve(Q_beta)
  l_beta <- t(Xtr_sc) %*% ytr + Omega_inv %*% xi
  
  # posterior draws
  sigma2_draws <- rinvgamma(
    samples,
    alpha0 + n_tr / 2,
    b0 + t(ytr) %*% ytr / 2 +
      t(xi) %*% Omega_inv %*% xi / 2 -
      t(l_beta) %*% Q_beta_inv %*% l_beta / 2
  )
  
  beta_draws <- matrix(NA, nrow = samples, ncol = p_tr)
  beta_mean_post <- as.vector(Q_beta_inv %*% l_beta)
  
  for (s in 1:samples) {
    beta_draws[s, ] <- mvrnorm(
      1, mu = beta_mean_post,
      Sigma = sigma2_draws[s] * Q_beta_inv
    )
  }
  
  beta_post_mean <- colMeans(beta_draws)
  
  # posterior mean predictions on test
  yhat_test <- as.vector(Xte_sc %*% beta_post_mean)
  
  # Test MSE
  mse <- mean((yte - yhat_test)^2)
  
  # test R^2
  sse <- sum((yte - yhat_test)^2)
  sst <- sum((yte - mean(yte))^2)
  r2  <- 1 - sse / sst
  
  list(r2 = r2, mse = mse, beta_post_mean = beta_post_mean)
}
```

### 1. Model 1: Full Model

```{r}
m1 <- fit_conjugate_blr(X_train_scaled, y_train,
                        X_test_scaled,  y_test)
```

### 2. Model 2: Bayesian Lasso

```{r}
set.seed(123)
bl_fit_tr <- blasso(
  X = X_train_scaled,
  y = y_train,
  T = 10000,
  thin = 5,
  verb = 0
)

beta_ci_tr <- t(apply(bl_fit_tr$beta, 2, quantile, probs = c(0.025, 0.975)))
colnames(beta_ci_tr) <- c("low", "high")  

selected_lasso_ci_tr <- rownames(beta_ci_tr)[
  beta_ci_tr[, "low"] > 0 | beta_ci_tr[, "high"] < 0
]

idx_sel_tr <- as.integer(sub("b\\.", "", selected_lasso_ci_tr))
selected_nutrients_tr <- colnames(X_train_scaled)[idx_sel_tr]

# subset train/test using the SAME columns
Xtr_BL_F <- X_train_scaled[, selected_nutrients_tr, drop = FALSE]
Xte_BL_F <- X_test_scaled[,  selected_nutrients_tr, drop = FALSE]

m2 <- fit_conjugate_blr(Xtr_BL_F, y_train,
                        Xte_BL_F, y_test)
```

### 3. Model 3: Bayesian Lasso Without Fat

```{r}
X_train_wofat <- X_train %>% dplyr::select(-Fat)
X_test_wofat  <- X_test  %>% dplyr::select(-Fat)

mu_wofat <- apply(X_train_wofat, 2, mean)
sd_wofat <- apply(X_train_wofat, 2, sd)

Xtr_wofat_sc <- scale_with_train(X_train_wofat, mu_wofat, sd_wofat)
Xte_wofat_sc <- scale_with_train(X_test_wofat,  mu_wofat, sd_wofat)

set.seed(123)
bl_fit_wofat_tr <- blasso(
  X = Xtr_wofat_sc,
  y = y_train,
  T = 10000,
  thin = 5,
  verb = 0
)

beta_ci_wofat_tr <- t(apply(bl_fit_wofat_tr$beta, 2, quantile, probs = c(0.025, 0.975)))
colnames(beta_ci_wofat_tr) <- c("low", "high")

selected_lasso_ci_wofat_tr <- rownames(beta_ci_wofat_tr)[
  beta_ci_wofat_tr[, "low"] > 0 | beta_ci_wofat_tr[, "high"] < 0
]

idx_sel_wofat_tr <- as.integer(sub("b\\.", "", selected_lasso_ci_wofat_tr))
selected_nutrients_wofat_tr <- colnames(Xtr_wofat_sc)[idx_sel_wofat_tr]

Xtr_BL <- Xtr_wofat_sc[, selected_nutrients_wofat_tr, drop = FALSE]
Xte_BL <- Xte_wofat_sc[, selected_nutrients_wofat_tr, drop = FALSE]

m3 <- fit_conjugate_blr(Xtr_BL, y_train,
                        Xte_BL, y_test)
```

### 4. Model 4: Bayesian Lasso Without Fat Types

```{r}
X_train_wofat_types <- X_train %>% dplyr::select(
  -c(Saturated_Fats, Monounsaturated_Fats, Polyunsaturated_Fats)
)
X_test_wofat_types <- X_test %>% dplyr::select(
  -c(Saturated_Fats, Monounsaturated_Fats, Polyunsaturated_Fats)
)

# Compute train mean/sd for correct scaling
mu_wofat_types <- apply(X_train_wofat_types, 2, mean)
sd_wofat_types <- apply(X_train_wofat_types, 2, sd)

# Apply scaling using training statistics
Xtr_wofat_types_sc <- scale_with_train(X_train_wofat_types, mu_wofat_types, sd_wofat_types)
Xte_wofat_types_sc <- scale_with_train(X_test_wofat_types,  mu_wofat_types, sd_wofat_types)

# Bayesian Lasso selection on TRAIN
set.seed(123)
bl_fit_wofat_types_tr <- blasso(
  X = Xtr_wofat_types_sc,
  y = y_train,
  T = 10000,
  thin = 5,
  verb = 0
)

# 95% CI for coefficients
beta_ci_wofat_types_tr <- t(apply(
  bl_fit_wofat_types_tr$beta,
  2,
  quantile,
  probs = c(0.025, 0.975)
))
colnames(beta_ci_wofat_types_tr) <- c("low", "high")

# Select coefficients whose 95% CI excludes zero
selected_lasso_ci_wofat_types_tr <- rownames(beta_ci_wofat_types_tr)[
  beta_ci_wofat_types_tr[, "low"] > 0 |
    beta_ci_wofat_types_tr[, "high"] < 0
]

# Map b.## to column names
idx_sel_wofat_types_tr <- as.integer(
  sub("b\\.", "", selected_lasso_ci_wofat_types_tr)
)
selected_nutrients_wofat_types_tr <- colnames(Xtr_wofat_types_sc)[idx_sel_wofat_types_tr]

# Final predictor matrices
Xtr_BL_nofat_types <- Xtr_wofat_types_sc[, selected_nutrients_wofat_types_tr, drop = FALSE]
Xte_BL_nofat_types <- Xte_wofat_types_sc[, selected_nutrients_wofat_types_tr, drop = FALSE]

# Fit conjugate BLR
m4 <- fit_conjugate_blr(Xtr_BL_nofat_types, y_train,
                        Xte_BL_nofat_types, y_test)
```

### 5. Model 5: Spike and Slab Prior

```{r}
set.seed(123)
df_model_tr <- data.frame(
  y = as.numeric(y_train),
  as.data.frame(X_train_scaled)
)

ss_fit_tr <- lm.spike(
  y ~ .,
  data  = df_model_tr,
  niter = 6000,
  expected.model.size = 10
)

burn_in <- 1000
ss_sum_tr <- summary(ss_fit_tr, burn = burn_in)
coef_table_tr <- ss_sum_tr$coefficients

inc_prob_tr <- coef_table_tr[, 5]
inc_prob_tr <- inc_prob_tr[names(inc_prob_tr) != "(Intercept)"]

selected_vars_tr <- names(inc_prob_tr)[inc_prob_tr > 0.5]

if (length(selected_vars_tr) == 0) {
  stop("Spike&Slab selected no variables on TRAIN (inc_prob > 0.5). Lower threshold.")
}

Xtr_SS <- X_train_scaled[, selected_vars_tr, drop = FALSE]
Xte_SS <- X_test_scaled[,  selected_vars_tr, drop = FALSE]

m5 <- fit_conjugate_blr(Xtr_SS, y_train,
                        Xte_SS, y_test)
```

### 6. Model 6: Spike and Slab Prior Without Fat

```{r}
X_train_wofat <- X_train %>% dplyr::select(-Fat)
X_test_wofat  <- X_test  %>% dplyr::select(-Fat)

mu_wofat <- apply(X_train_wofat, 2, mean)
sd_wofat <- apply(X_train_wofat, 2, sd)

Xtr_wofat_sc <- scale_with_train(X_train_wofat, mu_wofat, sd_wofat)
Xte_wofat_sc <- scale_with_train(X_test_wofat,  mu_wofat, sd_wofat)

set.seed(123)
df_model_tr_nofat <- data.frame(
  y = as.numeric(y_train),
  as.data.frame(Xtr_wofat_sc)
)

ss_fit_tr_nofat <- lm.spike(
  y ~ .,
  data  = df_model_tr_nofat,
  niter = 6000,
  expected.model.size = 10
)

burn_in <- 1000
ss_sum_tr_nofat <- summary(ss_fit_tr_nofat, burn = burn_in)
coef_table_tr_nofat <- ss_sum_tr_nofat$coefficients

inc_prob_tr_nofat <- coef_table_tr_nofat[, 5]
inc_prob_tr_nofat <- inc_prob_tr_nofat[names(inc_prob_tr_nofat) != "(Intercept)"]

selected_vars_tr_nofat <- names(inc_prob_tr_nofat)[inc_prob_tr_nofat > 0.5]

if (length(selected_vars_tr_nofat) == 0) {
  stop("Spike&Slab (No Fat) selected no variables on TRAIN (inc_prob > 0.5). Lower threshold.")
}

Xtr_SS_nofat <- Xtr_wofat_sc[, selected_vars_tr_nofat, drop = FALSE]
Xte_SS_nofat <- Xte_wofat_sc[,  selected_vars_tr_nofat, drop = FALSE]

m6 <- fit_conjugate_blr(Xtr_SS_nofat, y_train,
                        Xte_SS_nofat, y_test)
```

### 7. Model 7: Spike and Slab Prior Without Fat Types

```{r}
X_train_wofat_types <- X_train %>% dplyr::select(
  -c(Saturated_Fats, Monounsaturated_Fats, Polyunsaturated_Fats)
)
X_test_wofat_types <- X_test %>% dplyr::select(
  -c(Saturated_Fats, Monounsaturated_Fats, Polyunsaturated_Fats)
)

# Compute train mean/sd for correct scaling
mu_wofat_types <- apply(X_train_wofat_types, 2, mean)
sd_wofat_types <- apply(X_train_wofat_types, 2, sd)

# Apply scaling using training statistics
Xtr_wofat_types_sc <- scale_with_train(X_train_wofat_types, mu_wofat_types, sd_wofat_types)
Xte_wofat_types_sc <- scale_with_train(X_test_wofat_types,  mu_wofat_types, sd_wofat_types)

set.seed(123)
df_model_tr_nofat_types <- data.frame(
  y = as.numeric(y_train),
  as.data.frame(Xtr_wofat_types_sc)
)

ss_fit_tr_nofat_types <- lm.spike(
  y ~ .,
  data  = df_model_tr_nofat_types,
  niter = 6000,
  expected.model.size = 10
)

burn_in <- 1000
ss_sum_tr_nofat_types <- summary(ss_fit_tr_nofat_types, burn = burn_in)
coef_table_tr_nofat_types <- ss_sum_tr_nofat_types$coefficients

inc_prob_tr_nofat_types <- coef_table_tr_nofat_types[, 5]
inc_prob_tr_nofat_types <- inc_prob_tr_nofat_types[names(inc_prob_tr_nofat_types) != "(Intercept)"]

selected_vars_tr_nofat_types <- names(inc_prob_tr_nofat_types)[inc_prob_tr_nofat_types > 0.5]

if (length(selected_vars_tr_nofat_types) == 0) {
  stop("Spike&Slab (No Fat Types) selected no variables on TRAIN (inc_prob > 0.5). Lower threshold.")
}

Xtr_SS_nofat_types <- Xtr_wofat_types_sc[, selected_vars_tr_nofat_types, drop = FALSE]
Xte_SS_nofat_types <- Xte_wofat_types_sc[,  selected_vars_tr_nofat_types, drop = FALSE]

m7 <- fit_conjugate_blr(Xtr_SS_nofat_types, y_train,
                        Xte_SS_nofat_types, y_test)
```

### 8. Model 8: Fat+Carbs+Protein only

```{r}
vars_m8 <- c("Fat", "Carbohydrates", "Protein")

# make sure they exist
stopifnot(all(vars_m8 %in% colnames(X_train)))

# subset original (unscaled) train/test
X_train_m8 <- X_train[, vars_m8, drop = FALSE]
X_test_m8  <- X_test[,  vars_m8, drop = FALSE]

# scale using TRAIN stats of these 3 variables
mu_m8 <- apply(X_train_m8, 2, mean)
sd_m8 <- apply(X_train_m8, 2, sd)

Xtr_m8_sc <- scale_with_train(X_train_m8, mu_m8, sd_m8)
Xte_m8_sc <- scale_with_train(X_test_m8,  mu_m8, sd_m8)

m8 <- fit_conjugate_blr(Xtr_m8_sc, y_train,
                        Xte_m8_sc, y_test)
```

### Summary Table

```{r}
r2_table <- tibble(
  Model = c(
    "Model 1: Full",
    "Model 2: BLasso",
    "Model 3: BLasso (no Fat)",
    "Model 4: BLasso (no Fat Types)",
    "Model 5: Spike&Slab",
    "Model 6: Spike&Slab (no Fat)",
    "Model 7: Spike&Slab (no Fat Types)",
    "Model 8: Fat+Carbs+Protein only"
  ),
  Test_R2 = c(m1$r2, m2$r2, m3$r2, m4$r2, m5$r2, m6$r2, m7$r2, m8$r2),
  Test_MSE = c(m1$mse, m2$mse, m3$mse, m4$mse, m5$mse, m6$mse, m7$mse, m8$mse),
  Num_Predictors = c(
    ncol(X_train_scaled),      # Model 1
    ncol(Xtr_BL_F),            # Model 2
    ncol(Xtr_BL),              # Model 3
    ncol(Xtr_BL_nofat_types),  # Model 4
    ncol(Xtr_SS),              # Model 5
    ncol(Xtr_SS_nofat),        # Model 6
    ncol(Xtr_SS_nofat_types),  # Model 7
    ncol(Xtr_m8_sc)            # Model 8
  )
)
r2_table
```
